Descriptive, not injunctive, social norms caused increases in mask wearing throughout the COVID-19 pandemic

### 2022-10-17 manuscript.tex
second draft with edits

\linenumbers
\usepackage{csquotes}
\usepackage{array}
\usepackage{caption}
\usepackage{graphicx}
\usepackage{siunitx}
\usepackage[normalem]{ulem}
\usepackage{colortbl}
\usepackage{multirow}
\usepackage{hhline}
\usepackage{calc}
\usepackage{tabularx}
\usepackage{threeparttable}
\usepackage{wrapfig}
\usepackage{adjustbox}
\usepackage{hyperref}
\usepackage{setspace}
\raggedbottom
\AtBeginEnvironment{tabular}{\singlespacing}
\AtBeginEnvironment{lltable}{\singlespacing}
\AtBeginEnvironment{tablenotes}{\doublespacing}
\captionsetup[table]{font={stretch=1,small}}
\captionsetup[figure]{font={stretch=1,small}}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Descriptive, not injunctive, social norms caused increases in mask wearing throughout the COVID-19 pandemic},
  pdfauthor={Samantha L. Heiman,1, Scott Claessens,2, Edward R. Hurt1, \& Peter M. Todd1,3},
  pdflang={en-EN},
  pdfkeywords={descriptive norms; injunctive norms; longitudinal; COVID-19; mask wearing; cooperation},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{Descriptive, not injunctive, social norms caused increases in mask wearing throughout the COVID-19 pandemic}
\author{Samantha L. Heiman\textsuperscript{$\dagger{}$,1}, Scott Claessens\textsuperscript{$\dagger{}$,2}, Edward R. Hurt\textsuperscript{1}, \& Peter M. Todd\textsuperscript{1,3}}
\date{}

@@

\abstract{%
Social norms allow humans to coordinate and cooperate in the face of existential threats. In particular, injunctive social norms prescribe what people \emph{ought} to do, whereas descriptive social norms inform what people \emph{actually} do. While previous experimental work has revealed people's sensitivity to normative influence, several open questions remain about the natural emergence of injunctive and descriptive social norms within populations and their influences on cooperative behavior over time. To understand how social norms emerge and shape behavior in a non-experimental setting, we studied mask wearing during the COVID-19 pandemic. Leveraging two years of longitudinal data from a representative sample of adults in the United States (18 time points; \emph{n} = 915), we tracked people's reported mask wearing behavior and their perceived injunctive and descriptive mask wearing norms as the pandemic unfolded. Longitudinal trends of norm perceptions and self-reported mask wearing suggested that norms and behavior are tightly coupled and both change quickly in response to recommendations from public health authorities. In addition, a random-intercept cross-lagged panel model revealed that descriptive norms, but not injunctive norms, caused future increases in mask wearing behavior. These findings underline the relative importance of descriptive social norms in shaping cooperative behavior during uncertain times.
}

@@

Despite decades of research on the causes and consequences of injunctive and descriptive norms (1, 2, 16--18), social norms have remained an elusive concept in the behavioral sciences and several open questions remain (3, 5). In the current work, we focus on two such questions. First, how do injunctive and descriptive norms emerge naturally over time within a population? Second, how do evolving injunctive and descriptive norms influence behavior?

@@

Recent research has found positive relationships between perceptions of social norms and protective COVID-19 behaviors. In the United States, one cross-sectional study found that perceptions of injunctive norms positively predicted intentions to stay at home to minimize exposure (29), and another vignette study found that experimentally-manipulated descriptive norms increased personal mask wearing intentions (30). In Germany, a two-wave study found that perceptions of descriptive norms positively predicted future protective behaviors, such as physical distancing (31). These studies are telling, but since they are experimental, cross-sectional, or only minimally longitudinal, they are unable to distinguish between between-person and within-person change over time (32), nor do they have the temporal granularity to capture fluctuating changes in norm strength and norm adherence across the entire pandemic. These studies also do not control for the influences of non-social beliefs, such as factual beliefs and personal normative beliefs (27).

Here, we use two years of longitudinal data from a representative sample of adults in the United States (18 time points; \emph{n} = 916) to track the development of descriptive and injunctive mask wearing norms and mask wearing behavior over the course of the COVID-19 pandemic. We aimed to answer two main research questions. First, how do descriptive and injunctive mask wearing norms emerge and evolve over time in the United States population? Second, how do descriptive and injunctive mask wearing norms affect mask wearing behavior over time?

@@

\begin{figure}
\centering
\includegraphics{manuscript_files/figure-latex/plotTimeline-1.pdf}
\caption{\label{fig:plotTimeline}\emph{Timeline of self-reported mask wearing behavior and perceived social norms in the United States throughout the COVID-19 pandemic.} (a) Points and line ranges indicate mean averages and standard errors for the self-reported mask wearing item. This item was measured across all eighteen time points on a 5-point Likert scale, with higher values indicating increased frequency of personal mask wearing during in-person interactions. (b) Points and line ranges indicate mean averages and standard errors for perceived descriptive mask wearing norms (blue) and perceived injunctive mask wearing norms (green). These items were measured across eleven time points on a 7-point Likert scale, with higher values indicating stronger perceived social norms. (c) Smoothed data for new COVID-19 cases in the United States, displayed on the log scale (data retrieved from Our World in Data; \url{https://ourworldindata.org/}). Across all panels, gray dashed lines represent significant pandemic-related events in the United States, such as vaccine approval from the Food and Drug Administration (FDA) and public health recommendations from the Centers for Disease Control and Prevention (CDC).}
\end{figure}

Two main observations can be made about the emergence and stability of social norms from these visualizations. First, social norms and behavior were tightly coupled over time. Although social norms are measured on fewer occasions than mask wearing behavior, we can see that as mask wearing behavior decreased in the summer of 2021, so too did perceived descriptive and injunctive mask wearing norms. Subsequently, the steep rise in COVID-19 case numbers in the fall of 2021 saw concomitant increases in both mask wearing behavior and perceived social norms, though to lower levels than before. In line with these patterns, multilevel regression models revealed positive correlations between mask wearing behavior and perceived descriptive mask wearing norms (\emph{b} = 0.29, 95\% confidence interval {[}0.23 0.35{]}) and between mask wearing behavior and perceived injunctive mask wearing norms (\emph{b} = 0.26, 95\% CI {[}0.22 0.30{]}) across individuals and time points (Supplementary Figure \ref{fig:plotCorBehNorm}; Supplementary Table \ref{tab:modelSummaryTable1}).

Second, fluctuations in mask wearing behavior and perceived social norms are in line with recommendations broadcasted by the CDC, the main national public health agency of the United States. We do not have data for the very start of the pandemic in early 2020, but the high levels of mask wearing and strong perceived social norms at the start of our observation window likely emerged after the initial mask wearing recommendation from the CDC in April 2020. Perceived social norms and mask wearing behavior subsequently declined after the CDC rescinded their mask wearing recommendation in March 2021, and then increased again after the CDC updated their guidelines for indoor mask use in high risk areas in August 2021. Finally, perceived social norms and mask wearing declined again after the CDC eased restrictions in March 2022. Impressively, shifts in mask wearing behavior and norms happened on the timescale of weeks. These trends were confirmed by a series of multilevel regression models with change points aligning with changes in CDC mask wearing recommendations (Supplementary Figure \ref{fig:plotCDCSens}; Supplementary Table \ref{tab:changePointsTable}).

Sample averages can provide informative trends, but they do not allow us to estimate within-person changes in mask wearing behavior and perceived social norms over time. To determine whether within-person changes in social norms temporally preceded within-person changes in mask wearing behavior, we fitted a ten-wave unconstrained random-intercept cross-lagged panel model to the longitudinal data. This structural equation model separately estimated stable trait-like between-person individual differences and within-person fluctuations from trait levels for our main variables and controls: perceived descriptive mask wearing norms, perceived injunctive mask wearing norms, factual beliefs, personal normative beliefs, and self-reported mask wearing behavior (see Supplementary Figure \ref{fig:plotDAG} for justification of control variables). According to established fit statistics, this model fitted the data well (root mean square error of approximation = 0.030, 95\% CI {[}0.028 0.033{]}; standardized root mean squared residual = 0.087; comparative fit index = 0.957). Since we are primarily interested in the causal effects of social norms on behavior, we focus on these variables in what follows (but see Supplementary Table \ref{tab:lavaanTable} for full list of estimated autoregressive and cross-lagged effects).

@@

\begin{figure}
\centering
\includegraphics{manuscript_files/figure-latex/plotRICLPM-1.pdf}
\caption{\label{fig:plotRICLPM}\emph{Results of ten-wave unconstrained random-intercept cross-lagged panel model.} Arrows represent within-person autoregressive and cross-lagged effects from the model, partitioning out stable between-person individual differences and controlling for factual beliefs, personal normative beliefs, and political orientation. Arrow thickness is scaled according to standardized effect size. Bolded arrows indicate significantly positive parameters, \emph{p} \textless{} 0.05. Gray arrows indicate non-significant parameters.}
\end{figure}

In late 2020 and throughout 2021, we see several cross-lagged effects from perceived descriptive norms to future mask wearing behavior. On four occasions, within-person increases in perceived descriptive norms predicted future within-person increases in mask wearing behavior. According to recent effect size guidelines for cross-lagged panel models (34), the standardized beta coefficients for these cross-lagged effects were large (Time 2, \(\beta\) = 0.17, 95\% CI {[}0.06 0.28{]}, \emph{p} = .002; Time 5, \(\beta\) = 0.21, 95\% CI {[}0.08 0.34{]}, \emph{p} = .001; Time 11, \(\beta\) = 0.15, 95\% CI {[}0.01 0.30{]}, \emph{p} = .041; Time 13, \(\beta\) = 0.16, 95\% CI {[}0.02 0.29{]}, \emph{p} = .023). These cross-lagged effects from descriptive norms to mask wearing diminish in 2022. We also find some evidence for a reciprocal effect, whereby within-person increases in mask wearing behavior predicted future within-person increases in perceived descriptive norms. Moreover, several cross-lagged effects emerged between perceived descriptive and injunctive norms, demonstrating reciprocal within-person causal effects between these variables.

@@

In our longitudinal analysis, we found that \emph{descriptive} norms, not injunctive norms, independently predicted future increases in mask wearing. In line with this finding, descriptive norms have also been shown to predict future increases in physical distancing and prosocial behaviors (e.g., neighborhood help, charitable donations) throughout the COVID-19 pandemic, though this previous work did not adequately disentangle between-person and within-person effects (31). Similarly, experimental work has revealed that people in the United States are more likely to report intentions to wear a mask if they are told that others are wearing masks (30). Why this specific effect of descriptive norms on mask wearing? Descriptive norms are particularly useful for fast changing, threatening situations with a high degree of uncertainty, such as the COVID-19 pandemic (36). During times of uncertainty, people look to others to quickly coordinate their behavior, and attempt to alleviate uncertainty-related stress by identifying with their group and its social norms (37, 38). Supporting this situational-uncertainty explanation, our model revealed that descriptive norms predicted future mask wearing at the beginning of the pandemic, when uncertainty was likely at its height. Unsure of how to respond to a novel threatening situation, people began to look to their neighbors and adapted their mask wearing behavior accordingly. However, as the uncertainty surrounding the pandemic lessened in 2022, reliance on descriptive norms appeared to decrease.

Our finding that injunctive norms do not predict future mask wearing behavior is at odds with cross-sectional evidence showing that perceived injunctive norms are positively correlated with intentions to stay indoors during the pandemic among older adults (29). One possible explanation for these conflicting findings is that, unlike previous work, our model systematically controlled for non-social beliefs, such as factual beliefs and personal normative beliefs, which could potentially have driven previous correlations. Another explanation is that, due to the increased opportunities to observe mask wearing in public, descriptive norms of mask wearing behavior were made more salient than injunctive norms throughout the pandemic, and therefore had a greater influence on behavior (2). By contrast, for more private behaviors like remaining indoors, it would have been less possible to observe other people's behaviors, increasing the relative salience of injunctive norms. To test this idea, future research should expand our longitudinal cross-lagged approach to protective behaviors beyond mask wearing, including both public behaviors (e.g., physical distancing) and private behaviors (e.g., hand washing and home isolation).

@@

\hypertarget{methods}{%
\section{Methods}\label{methods}}

@@

To analyze average trends in self-reported mask wearing behavior and perceived social norms, we fitted several multilevel regression models. First, to determine whether mask wearing behavior and social norms were coupled over time, we regressed mask wearing behavior on perceived descriptive and injunctive norms separately, including random intercepts and slopes for participants and time points. Second, to analyze changes over time, we regressed mask wearing behavior and perceived social norms onto a continuous time predictor. These models included random intercepts and slopes for participants, as well as change points aligning with changes in CDC mask wearing recommendations. We estimated these models using the \emph{lme4} R package (43) and dealt with missing data via listwise deletion.

@@

We estimated the random-intercept cross-lagged panel model using the \emph{lavaan} R package (45). In line with our directed acyclic graph (see Supplementary Figure \ref{fig:plotDAG}), we included five main variables in the model: perceived descriptive norms, perceived injunctive norms, factual beliefs, personal normative beliefs, and self-reported mask wearing behavior. For each of these variables, the model estimated a stable between-person trait level (random intercept) and time-specific within-person fluctuations from this trait level. We modelled autoregressive and cross-lagged effects between all five variables, and included political orientation as a time-invariant covariate. We restricted the analysis to the ten time points with available data for all five variables. Full information maximum likelihood estimation was used to deal with missing data.

@@

\leavevmode\vadjust pre{\hypertarget{ref-Asch1956}{}}%
\CSLLeftMargin{6. }%
\CSLRightInline{S. E. Asch, \href{https://doi.org/10.1037/h0093718}{Studies of independence and conformity: I. A minority of one against a unanimous majority}. \emph{Psychological monographs: General and applied} \textbf{70}, 1 (1956).}

@@

\leavevmode\vadjust pre{\hypertarget{ref-Deutsch1955}{}}%
\CSLLeftMargin{15. }%
\CSLRightInline{M. Deutsch, H. B. Gerard, \href{https://doi.org/10.1037/h0046408}{A study of normative and informational social influences upon individual judgment}. \emph{The journal of abnormal and social psychology} \textbf{51}, 629--636 (1955).}

@@

\leavevmode\vadjust pre{\hypertarget{ref-Boyd1985}{}}%
\CSLLeftMargin{19. }%
\CSLRightInline{\&. R. Boyd Robert, \emph{Culture and the evolutionary process} (University of Chicago Press, 1985).}

@@

\leavevmode\vadjust pre{\hypertarget{ref-Jordan2009}{}}%
\CSLLeftMargin{21. }%
\CSLRightInline{F. M. Jordan, R. D. Gray, S. J. Greenhill, R. Mace, \href{https://doi.org/10.1098/rspb.2009.0088}{Matrilocal residence is ancestral in austronesian societies}. \emph{Proceedings of the Royal Society B: Biological Sciences} \textbf{276}, 1957--1964 (2009).}

@@

\leavevmode\vadjust pre{\hypertarget{ref-Hamaker2015}{}}%
\CSLLeftMargin{32. }%
\CSLRightInline{E. L. Hamaker, R. M. Kuiper, R. P. P. P. Grasman, \href{https://doi.org/10.1037/a0038889}{A critique of the cross-lagged panel model}. \emph{Psychological methods} \textbf{20}, 102 (2015).}

@@

To evaluate the construct validity of our measures of perceived descriptive and injunctive norms, at Time 7 we asked participants to rate the extent to which each perceived norm item provided descriptive and injunctive information. For each item, participants were asked whether the item provided information about what people are doing, and whether the item provided information about what people should be doing. Participants responded on a 7-point Likert scale, from (1) Not At All to (7) Very Strongly. For a full list of questions, see Supplementary Table \ref{tab:itemTable}.

@@

@@

\begin{figure}
\centering
\includegraphics{manuscript_files/figure-latex/plotCorBehNorm-1.pdf}
\caption{\label{fig:plotCorBehNorm}\emph{Predictions from multilevel models with self-reported mask wearing behavior as the outcome variable and (a) perceived strength of descriptive norms and (b) perceived strength of injunctive norms as independent predictor variables.} Lines are fixed effect regression lines from multilevel models, shaded areas are 95\% confidence intervals.}
\end{figure}

\begin{figure}
\centering
\includegraphics{manuscript_files/figure-latex/plotCDCSens-1.pdf}
\caption{\label{fig:plotCDCSens}\emph{Predictions from multilevel models with change points in line with changes in CDC mask wearing recommendations.} These models track temporal changes in (a) self-reported mask wearing behavior, (b) perceived strength of descriptive norms, and (c) perceived strength of injunctive norms. Time is included as a continuous linear predictor, scaled between 0 and 1, with two forced change points (dashed lines). The first dashed line (left) indicates when the CDC relaxed their mask wearing recommendations in March 2021, and the second dashed line (right) indicates when the CDC strengthened their mask wearing recommendations in July 2021. This results in the estimation of four fixed effect parameters: the initial intercept, the slope in the first window, the slope in the second window, and the slope in the third window. Bolded lines and shaded areas represent fixed effect regression lines from multilevel models and 95\% confidence intervals, respectively.}
\end{figure}

>>>>

\linenumbers
\usepackage{csquotes}
\usepackage{array}
\usepackage{caption}
\usepackage{graphicx}
\usepackage{siunitx}
\usepackage[normalem]{ulem}
\usepackage{colortbl}
\usepackage{multirow}
\usepackage{hhline}
\usepackage{calc}
\usepackage{tabularx}
\usepackage{threeparttable}
\usepackage{wrapfig}
\usepackage{adjustbox}
\usepackage{hyperref}
\usepackage{setspace}
\raggedbottom
\AtBeginEnvironment{tabular}{\singlespacing}
\AtBeginEnvironment{lltable}{\singlespacing}
\AtBeginEnvironment{tablenotes}{\doublespacing}
\captionsetup[table]{font={stretch=1,small}}
\captionsetup[figure]{font={stretch=1,small}}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Descriptive, not injunctive, social norms caused increases in mask wearing throughout the COVID-19 pandemic},
  pdfauthor={Samantha L. Heiman,1, Scott Claessens,2, Edward R. Hirt1, \& Peter M. Todd1,3},
  pdflang={en-EN},
  pdfkeywords={descriptive norms; injunctive norms; longitudinal; COVID-19; mask wearing; cooperation},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{Descriptive, not injunctive, social norms caused increases in mask wearing throughout the COVID-19 pandemic}
\author{Samantha L. Heiman\textsuperscript{$\dagger{}$,1}, Scott Claessens\textsuperscript{$\dagger{}$,2}, Edward R. Hirt\textsuperscript{1}, \& Peter M. Todd\textsuperscript{1,3}}
\date{}

@@

\abstract{%
Social norms allow humans to coordinate and cooperate in the face of existential threats. In particular, injunctive social norms prescribe what people \emph{ought} to do, whereas descriptive social norms inform what people \emph{actually} do. While previous experimental work has revealed people's sensitivity to normative influence, several open questions remain about the natural emergence of injunctive and descriptive social norms within populations and their causal influences on cooperative behavior over time. To understand how social norms emerge and shape behavior in a non-experimental setting, we studied mask wearing during the COVID-19 pandemic. Leveraging two years of longitudinal data from a representative sample of adults in the United States (18 time points; \emph{n} = 915), we tracked people's reported mask wearing behavior and their perceived injunctive and descriptive mask wearing norms as the pandemic unfolded. Longitudinal trends of norm perceptions and self-reported mask wearing suggested that norms and behavior were tightly coupled and both changed quickly in response to recommendations from public health authorities. In addition, a random-intercept cross-lagged panel model revealed that descriptive norms, but not injunctive norms, caused future increases in mask wearing behavior. These findings underline the relative importance of descriptive social norms in shaping cooperative behavior during uncertain times.
}

@@

Despite decades of research on the causes and consequences of injunctive and descriptive norms (1, 2, 16--18), social norms have remained an elusive concept in the behavioral sciences and several open questions remain (3, 5). In the current work, we focus on two such questions. First, how do injunctive and descriptive norms emerge naturally over time within a population? Second, how do evolving injunctive and descriptive norms causally influence behavior over time?

@@

Recent research has found positive relationships between perceptions of social norms and protective COVID-19 behaviors. In the United States, one cross-sectional study found that perceptions of injunctive norms positively predicted intentions to stay at home to minimize exposure (29), and another vignette study found that experimentally-manipulated descriptive norms increased personal mask wearing intentions (30). In Germany, a two-wave study found that perceptions of descriptive norms positively predicted future protective behaviors, such as physical distancing (31). These studies are telling, but since they are experimental, cross-sectional, or only minimally longitudinal, they are unable to distinguish between within-person and between-person change over time (32), nor do they have the temporal granularity to capture fluctuating changes in norm strength and norm adherence across the entire pandemic. These studies also do not adequately distinguish between the effects of social norms and the effects of other non-social beliefs that are also predicted to shape behavior, such as factual beliefs and personal normative beliefs (27).

Here, we use two years of longitudinal data from a representative sample of adults in the United States (18 time points; \emph{n} = 916) to track the development of descriptive and injunctive mask wearing norms and mask wearing behavior over the course of the COVID-19 pandemic. Between September 2020 and October 2022, we asked participants to report their frequency of mask wearing during in-person interactions, as well as their perceptions of descriptive and injunctive mask wearing norms. We used these data to answer two main research questions. First, how do descriptive and injunctive mask wearing norms emerge and evolve over time in the United States population? Second, how do descriptive and injunctive mask wearing norms causally influence mask wearing behavior over time?

@@

\begin{figure}
\centering
\includegraphics{manuscript_files/figure-latex/plotTimeline-1.pdf}
\caption{\label{fig:plotTimeline}\emph{Timeline of self-reported mask wearing behavior and perceived social norms in the United States throughout the COVID-19 pandemic.} (a) Points and line ranges indicate mean averages and standard errors for the self-reported mask wearing item. This item was measured across all eighteen time points on a 5-point Likert scale, with higher values indicating increased frequency of personal mask wearing during in-person interactions. (b) Points and line ranges indicate mean averages and standard errors for perceived injunctive mask wearing norms (green) and perceived descriptive mask wearing norms (blue). These items were measured across eleven time points on a 7-point Likert scale, with higher values indicating stronger perceived social norms. (c) Smoothed data for new COVID-19 cases in the United States, displayed on the log scale (data retrieved from Our World in Data; \url{https://ourworldindata.org/}). Across all panels, gray dashed lines represent significant pandemic-related events in the United States, such as vaccine approval from the Food and Drug Administration (FDA) and public health recommendations from the Centers for Disease Control and Prevention (CDC).}
\end{figure}

Two main observations can be made about the emergence and stability of social norms from these visualizations. First, social norms and behavior were tightly coupled over time. Although social norms are measured on fewer occasions than mask wearing behavior, we can see that as mask wearing behavior decreased in the summer of 2021, so too did perceived descriptive and injunctive mask wearing norms. Subsequently, the steep rise in COVID-19 case numbers in the fall of 2021 saw concomitant increases in both mask wearing behavior and perceived social norms, before declining again in 2022. In line with these patterns, multilevel regression models revealed positive correlations between mask wearing behavior and perceived descriptive mask wearing norms (\emph{b} = 0.29, 95\% confidence interval {[}0.23 0.35{]}, \emph{p} \textless{} .001) and between mask wearing behavior and perceived injunctive mask wearing norms (\emph{b} = 0.26, 95\% CI {[}0.22 0.30{]}, \emph{p} \textless{} .001) across individuals and time points (Supplementary Figure \ref{fig:plotCorBehNorm}; Supplementary Table \ref{tab:modelSummaryTable1}).

Second, fluctuations in mask wearing behavior and perceived social norms are in line with recommendations broadcasted by the CDC, the main national public health agency of the United States. We do not have data for the very start of the pandemic in early 2020, but the high levels of mask wearing and strong perceived social norms at the start of our observation window likely emerged after the initial mask wearing recommendation from the CDC in April 2020. Perceived social norms and mask wearing behavior subsequently declined after the CDC rescinded their mask wearing recommendation in March 2021, and then increased again after the CDC updated their guidelines for indoor mask use in high risk areas in August 2021. Finally, perceived social norms and mask wearing declined again after the CDC eased mask wearing guidelines in March 2022. Impressively, shifts in mask wearing behavior and norms happened on the timescale of weeks. These trends were confirmed by a series of multilevel regression models with change points aligning with changes in CDC mask wearing recommendations (Supplementary Figure \ref{fig:plotCDCSens}; Supplementary Table \ref{tab:changePointsTable}).

Sample averages can provide informative trends, but they do not allow us to estimate within-person changes in mask wearing behavior and perceived social norms over time. To determine whether within-person changes in social norms temporally preceded within-person changes in mask wearing behavior, we fitted a ten-wave unconstrained random-intercept cross-lagged panel model to the longitudinal data. This structural equation model separately estimated stable trait-like between-person individual differences and within-person fluctuations from trait levels for our main variables (self-reported mask wearing behavior, perceived descriptive mask wearing norms, and perceived injunctive mask wearing norms) and time-variant control variables (factual beliefs and personal normative beliefs). In line with our proposed causal model (Supplementary Figure \ref{fig:plotDAG}), we also included political orientation as an exogenous time-invariant control. According to established fit statistics, this model fitted the data well (root mean square error of approximation = 0.030, 95\% CI {[}0.028 0.033{]}; standardized root mean squared residual = 0.087; comparative fit index = 0.957). Since we are primarily interested in the causal effects of social norms on behavior, in what follows we focus on the results for mask wearing behavior, perceived descriptive norms, and perceived injunctive norms (but see Supplementary Table \ref{tab:lavaanTable} for full list of estimated autoregressive and cross-lagged effects).

@@

\begin{figure}
\centering
\includegraphics{manuscript_files/figure-latex/plotRICLPM-1.pdf}
\caption{\label{fig:plotRICLPM}\emph{Results of ten-wave unconstrained random-intercept cross-lagged panel model.} Arrows represent within-person autoregressive and cross-lagged effects for mask wearing behavior and perceived descriptive and injunctive norms, partitioning out stable between-person individual differences and controlling for factual beliefs, personal normative beliefs, and political orientation. Arrow thickness is scaled according to standardized effect size. Bolded arrows indicate significantly positive parameters, \emph{p} \textless{} 0.05. Gray arrows indicate non-significant parameters.}
\end{figure}

In late 2020 and throughout 2021, we see several cross-lagged effects from perceived descriptive norms to future mask wearing behavior. On four occasions, within-person increases in perceived descriptive norms predicted future within-person increases in mask wearing behavior. According to recent effect size guidelines for cross-lagged panel models (34), the standardized beta coefficients for these cross-lagged effects were large (Time 2, \(\beta\) = 0.17, 95\% CI {[}0.06 0.28{]}, \emph{p} = .002; Time 5, \(\beta\) = 0.21, 95\% CI {[}0.08 0.34{]}, \emph{p} = .001; Time 11, \(\beta\) = 0.15, 95\% CI {[}0.01 0.30{]}, \emph{p} = .041; Time 13, \(\beta\) = 0.16, 95\% CI {[}0.02 0.29{]}, \emph{p} = .023). In 2022, these cross-lagged effects from descriptive norms to mask wearing behavior diminished. We also find some evidence for a reciprocal effect, whereby within-person increases in mask wearing behavior predicted future within-person increases in perceived descriptive norms. Moreover, several cross-lagged effects emerged between perceived descriptive and injunctive norms, demonstrating reciprocal within-person causal effects between these variables.

@@

In our longitudinal analysis, we found that descriptive norms, not injunctive norms, predicted future increases in mask wearing across multiple time points. These cross-lagged effects were independent of the effects of non-social beliefs and political orientation. In line with this finding, descriptive norms have also been shown to predict future increases in physical distancing and prosocial behaviors (e.g., neighborhood help, charitable donations) throughout the COVID-19 pandemic, though this previous work did not adequately disentangle between-person and within-person effects (31). Similarly, experimental work has revealed that people in the United States are more likely to report intentions to wear a mask if they are told that others are wearing masks (30). Why this specific effect of descriptive norms on mask wearing? Descriptive norms are particularly useful for fast changing, threatening situations with a high degree of uncertainty, such as the COVID-19 pandemic (36). During times of uncertainty, people look to others to quickly coordinate their behavior, and attempt to alleviate uncertainty-related stress by identifying with their group and its social norms (37, 38). Supporting this situational-uncertainty explanation, our model revealed that descriptive norms predicted future mask wearing at the beginning of the pandemic, when uncertainty was likely at its peak. Unsure of how to respond to a novel threatening situation, people began to look to their neighbors and adapted their mask wearing behavior accordingly. However, as the uncertainty surrounding the pandemic lessened in 2022, reliance on descriptive norms appeared to decrease.

Our finding that injunctive norms do not predict future mask wearing behavior is at odds with cross-sectional evidence showing that perceived injunctive norms are positively correlated with intentions to stay indoors during the pandemic among older adults (29). One possible explanation for these conflicting findings is that, unlike previous work, our model systematically controlled for non-social beliefs, such as factual beliefs and personal normative beliefs, which could potentially have driven previous correlations. Another possible explanation is that, due to the increased opportunities to observe mask wearing in public, descriptive norms of mask wearing behavior were made more salient than injunctive norms throughout the pandemic, and therefore had a greater influence on behavior (2). By contrast, for more private behaviors like remaining indoors, it would have been less possible to observe other people's behaviors, increasing the relative salience of injunctive norms. To test this idea, future research should expand our longitudinal cross-lagged approach to protective behaviors beyond mask wearing, including both public behaviors (e.g., physical distancing) and private behaviors (e.g., hand washing and home isolation).

@@

\hypertarget{materials-and-methods}{%
\section{Materials and Methods}\label{materials-and-methods}}

@@

To analyze average trends in self-reported mask wearing behavior and perceived social norms, we fitted several multilevel regression models. First, to determine whether mask wearing behavior and social norms were coupled over time, we regressed mask wearing behavior on perceived descriptive and injunctive norms separately, including random intercepts and slopes for participants and time points. Second, to analyze whether changes over time were related to recommendations from the CDC, we regressed mask wearing behavior and perceived social norms onto a continuous time predictor. These models included random intercepts and slopes for participants, as well as change points aligning with changes in CDC mask wearing recommendations. We estimated these multilevel regression models using the \emph{lme4} R package (43) and dealt with missing data via listwise deletion.

@@

We estimated the random-intercept cross-lagged panel model using the \emph{lavaan} R package (45). In line with our directed acyclic graph (see Supplementary Figure \ref{fig:plotDAG}), we included three main variables (self-reported mask wearing behavior, perceived descriptive norms, and perceived injunctive norms) and two time-variant control variables (factual beliefs and personal normative beliefs) in the model. For each of these variables, the model estimated a stable between-person trait level (random intercept) and time-specific within-person fluctuations from this trait level. We modelled autoregressive and cross-lagged effects between all five variables, and included political orientation as a time-invariant covariate. We restricted the analysis to the ten time points with available data for all five variables. Full information maximum likelihood estimation was used to deal with missing data.

@@

\leavevmode\vadjust pre{\hypertarget{ref-Asch1956}{}}%
\CSLLeftMargin{6. }%
\CSLRightInline{S. E. Asch, \href{https://doi.org/10.1037/h0093718}{Studies of independence and conformity: I. A minority of one against a unanimous majority}. \emph{Psychological Monographs: General and Applied} \textbf{70}, 1 (1956).}

@@

\leavevmode\vadjust pre{\hypertarget{ref-Deutsch1955}{}}%
\CSLLeftMargin{15. }%
\CSLRightInline{M. Deutsch, H. B. Gerard, \href{https://doi.org/10.1037/h0046408}{A study of normative and informational social influences upon individual judgment}. \emph{The Journal of Abnormal and Social Psychology} \textbf{51}, 629--636 (1955).}

@@

\leavevmode\vadjust pre{\hypertarget{ref-Boyd1985}{}}%
\CSLLeftMargin{19. }%
\CSLRightInline{R. Boyd, P. J. Richerson, \emph{Culture and the evolutionary process} (University of Chicago Press, 1985).}

@@

\leavevmode\vadjust pre{\hypertarget{ref-Jordan2009}{}}%
\CSLLeftMargin{21. }%
\CSLRightInline{F. M. Jordan, R. D. Gray, S. J. Greenhill, R. Mace, \href{https://doi.org/10.1098/rspb.2009.0088}{Matrilocal residence is ancestral in {A}ustronesian societies}. \emph{Proceedings of the Royal Society B: Biological Sciences} \textbf{276}, 1957--1964 (2009).}

@@

\leavevmode\vadjust pre{\hypertarget{ref-Hamaker2015}{}}%
\CSLLeftMargin{32. }%
\CSLRightInline{E. L. Hamaker, R. M. Kuiper, R. P. P. P. Grasman, \href{https://doi.org/10.1037/a0038889}{A critique of the cross-lagged panel model}. \emph{Psychological Methods} \textbf{20}, 102 (2015).}

@@

To evaluate the construct validity of our measures of perceived descriptive and injunctive norms, at Time 7 we asked participants to rate the extent to which each perceived norm item provided descriptive and injunctive information. For each item, participants were asked whether the item provided information about what people \emph{are} doing, and whether the item provided information about what people \emph{should} be doing. Participants responded on a 7-point Likert scale, from (1) Not At All to (7) Very Strongly. For a full list of questions, see Supplementary Table \ref{tab:itemTable}.

@@

\begin{figure}
\centering
\includegraphics{manuscript_files/figure-latex/plotCorBehNorm-1.pdf}
\caption{\label{fig:plotCorBehNorm}\emph{Predictions from multilevel models with self-reported mask wearing behavior as the outcome variable and (a) perceived strength of descriptive norms and (b) perceived strength of injunctive norms as independent predictor variables.} Models contain random intercepts for participants and time points. Lines are fixed effect regression lines from multilevel models, shaded areas are 95\% confidence intervals.}
\end{figure}

\begin{figure}
\centering
\includegraphics{manuscript_files/figure-latex/plotCDCSens-1.pdf}
\caption{\label{fig:plotCDCSens}\emph{Predictions from multilevel models with change points in line with changes in CDC mask wearing recommendations.} These models track temporal changes in (a) self-reported mask wearing behavior, (b) perceived strength of descriptive norms, and (c) perceived strength of injunctive norms. Time is included as a continuous linear predictor, scaled between 0 and 1, with three forced change points (dashed lines). From the left, the first dashed line indicates when the CDC relaxed their mask wearing recommendations in March 2021, the second dashed line indicates when the CDC strengthened their mask wearing recommendations in July 2021, and the third dashed line indicates when the CDC updated their community levels and relaxed their mask wearing recommendations in March 2022. This results in the estimation of five fixed effect parameters: the initial intercept, the slope in the first window, the slope in the second window, the slope in the third window, and the slope is the fourth window. Bolded lines and shaded areas represent fixed effect regression lines from multilevel models and 95\% confidence intervals, respectively.}
\end{figure}

@@
